## Naive Bayes

Naive Bayes classifier is a supervised learning algorithm based on **Bayes' 
Theorem** for solving classification problems [@domingos1997optimal]. This 
method determines which category a sample belongs to by calculating the 
posterior probability of each category, and assumes that each feature is 
independent of each other under the conditions of the given category.

This "naive" assumption greatly simplifies the model computation and makes 
naive Bayes a classical algorithm that is both fast and stable.

In statistical learning and machine learning, naive Bayes is widely used for 
tasks such as text classification, medical diagnosis, credit risk assessment, 
and anomaly detection. Despite its relatively strict assumptions, it still 
performs well in high dimensional data scenarios due to its robustness, low 
computational cost, and good interpretability [@rish2001empirical].

### Theoretical basis

#### Bayes theorem

The theoretical basis of Naive Bayes classifier is **Bayes' Theorem**. It 
describes how to update our level of trust in an event after new observation 
information is available.

##### Basic form

The basic form of Bayes theorem is:

$$
P(A|B) = \frac{P(B|A)\,P(A)}{P(B)}.
$$

Among them:

-   $P(A)$: **Prior Probability** of event $A$, representing the initial belief 
about event $A$ in the absence of any additional information.
-   $P(B|A)$: **Likelihood**, denotes the probability of observing $B$ given 
the occurrence of $A$.
-   $P(A|B)$: **Posterior Probability**, which is the updated probability of 
event $A$ occurring after the observation of $B$.
-   $P(B)$: **Marginal Probability** of event $B$, which represents the overall 
probability of observing $B$ under all possible scenarios.

##### Application to classification problems

In the classification task, we let:

-   $Y$ is the category variable (e.g., "Is it a serious accident", "is it 
spam");
-   $X = (x_1, x_2, \ldots, x_n)$ is the observed eigenvector.

Bayes theorem can be rewritten as follows.

$$
P(Y|X) = \frac{P(X|Y)\,P(Y)}{P(X)}.
$$

Here:

-   $P(Y)$: prior probability, reflecting the distribution of each category in 
the population sample;
-   $P(X|Y)$: likelihood function, which represents the probability of 
occurrence of feature $X$ under the category $Y$;
-   $P(X)$: the normalization term, which is used to ensure that the posterior 
probabilities of all categories sum to one;
-   $P(Y|X)$: posterior probability, which represents the likelihood that the 
sample after a given feature $X$ belongs to the class $Y$.

Because $P(X)$ is the same for all categories, it can be ignored when comparing 
which category has a greater probability. The prediction criteria of the model 
are therefore simplified as follows.

$$
\hat{Y} = \arg\max_Y P(Y|X) = \arg\max_Y P(X|Y)\,P(Y).
$$

In other words: Instead of calculating all probabilities, we can simply compare 
the "prior × likelihood" of each category which is greater.

##### Intuitive explanation

Bayes' theorem can be understood as a "belief update formula" : When we obtain 
new observational information $X$, our belief about a certain category $Y$ 
changes.

| stage | concept | meaning |
|------------------------|------------------------|------------------------|
| Prior $P(Y)$ | Our original knowledge of the category $Y$ | For example: 30% <br> of serious accidents |
| Likelihood $P(X|Y)$ | probability of feature $X$ under category $Y$ | For example: a higher proportion of <br> serious accidents in nighttime accidents |
| A posterior $P(Y|X)$ | update probability after combining observations | For example, if it is at night, <br> the probability of a serious accident rises to about 56% |


In other words, Bayes' theorem lets us **update beliefs** with data. It is this 
updating mechanism that naive Bayes uses for classification.

#### Conditional independence assumption

Naive Bayes' "naive" comes from a key assumption: Given the category $Y$, the 
individual features $x_i$ are **conditionally independent** among each other 
[@domingos1997optimal]. Namely:

$$
P(X|Y) = \prod_{i=1}^{n} P(x_i|Y)
$$

This assumption greatly simplifies the calculation of the joint distribution. 
Originally we needed to estimate $P(x_1,x_2,... x_n|Y)$ such a high-dimensional 
joint probability, But with the independence assumption, one only needs to 
estimate the conditional probability $P(x_i|Y)$ of each individual feature.

This is not always true in real data, but in most scenarios it is still 
satisfactory, This is especially the case when the correlation between features 
is weak or the sample size is large enough.

#### Parameter estimates with likelihood functions

The core of model training is the estimation of two types of parameters:

I.  **Prior probability** $P(Y)$

Represents the proportion of categories occurring in the population:

$$
    \hat{P}(Y_k) = \frac{N_k}{N},
$$

Where $N_k$ is the number of samples for category $k$ and $N$ is the total 
number of samples.

II. **Conditional probability** $P(x_i|Y)$

-   If the feature is discrete:

$$
    \hat{P}(x_i|Y) = \frac{\text{count}(x_i,Y)}{\text{count}(Y)}.
$$

-   If the feature is continuous, assume that it follows a Gaussian 
distribution under each category:

$$
    P(x_i|Y) = \frac{1}{\sqrt{2\pi\sigma_{Y,i}^2}}
    \exp\! \left(-\frac{(x_i-\mu_{Y,i})^2}{2\sigma_{Y,i}^2}\right)
$$

where $\mu_{Y,i}$ and $\sigma_{Y,i}^2$ are the mean and variance of feature 
$x_i$ among all samples in class $Y$, respectively.

These parameters are usually obtained by Maximum Likelihood Estimation (MLE).

#### Class Prediction and Log Form

The goal of model prediction is to find the class that maximizes the posterior 
probability:

$$
\hat{Y} = \arg\max_Y P(Y) \prod_{i=1}^n P(x_i|Y)
$$

However, in practice, because continuous multiplication can lead to numerical 
underflow, We usually take logarithms and turn multiplication into addition:

$$
\log P(Y|X) = \log P(Y) + \sum_{i=1}^{n} \log P(x_i|Y)
$$

The final output of the model is:

$$
\hat{Y} = \arg\max_Y \big[\log P(Y) + \sum_i \log P(x_i|Y)\big].
$$

In this way, not only accuracy problems can be avoided, but also the 
contribution of each feature to the classification result can be observed more 
intuitively.

#### Theoretical Performance and Advantages and Limitations

I.  **Advantages**

-   The algorithm is simple, the computational complexity is linear, and the 
training and prediction speed is extremely fast;
-   Good performance on high dimensional data (such as bag-of-words model);
-   Good performance on small sample data; The model results are highly 
interpretable; Certain robustness against noisy data and missing data.

II. **Limitations**

Conditional independence assumption is often not true, and performance may 
decrease when feature correlation is strong; Inability to capture feature 
interaction effects; For the values of features that have not appeared before, 
there may be a zero probability problem; Output probability estimates are not 
necessarily reliable (often biased toward 0 or 1).

#### Laplace Smoothing

In naive Bayes, if a feature value never occurs in the training set, its 
corresponding probability is estimated to be zero. As a result, the whole 
product is 0 and the classifier outputs the wrong result. To avoid this 
situation, **Laplace Smoothing** is introduced.

That is to add a constant $\alpha > 0$ (usually 1) to the probability estimate:

$$
\hat{P}(x_i|Y) =
\frac{\text{count}(x_i,Y) + \alpha}
{\sum_{x_i'}(\text{count}(x_i',Y) + \alpha)}.
$$

So even if a value doesn't appear in the training set, the probability is not 
zero, The model is thus more stable and robust.

#### Surprisingly Good Performance

Although the conditional independence assumption is often not strictly true, 
naive Bayes still performs surprisingly well in practice. This is mainly due to 
the following:

-   **Robustness** : Even if there is some correlation between features, the 
classification result is still accurate as long as the dependence between 
different categories is not very different.

-   **Error Compensation** : The bias caused by conditional independence 
assumption often cancerates each other when multiple features are combined.

-   **Implicit Regularization** : Feature independence makes the model 
complexity low and can effectively prevent overfitting.

-   **High-dimensional Advantage** : In scenarios where the number of features 
is much larger than the number of samples, such as text classification, naive 
Bayes is almost always more stable than linear models.

### Types of Naive Bayes

There is not just one form of naive Bayesian algorithm. According to the type 
of feature variable (continuous or discrete), the value characteristics of the 
feature (Boolean, count, frequency, etc.) and the distribution assumption of 
the data, naive Bayes can be divided into the following main types:

#### Gaussian Naive Bayes

When the data feature is **continuous variable**, each feature is assumed to 
follow **Gaussian distribution** under a given category:

$$
P(x_i|Y) = \frac{1}{\sqrt{2\pi\sigma_{Y,i}^2}}
\exp\! \left(-\frac{(x_i - \mu_{Y,i})^2}{2\sigma_{Y,i}^2}\right)
$$

The advantages of Gaussian naive Bayes are:

-   suitable for continuous characteristics (e.g. age, temperature, price, 
etc.).
-   Computationally efficient.
-   Parameter estimation only requires sample mean and variance.
-   Good results can still be achieved for nonlinear separable data.

#### Categorical Naive Bayes

Categorical naive Bayes is used when the feature is **discrete (categorical) 
variable**. Assume that each feature follows a **categorical distribution** 
under the category $Y$:

$$
P(x_i|Y) = \frac{\text{count}(x_i,Y) + \alpha}
{\sum_{x_i'} (\text{count}(x_i',Y) + \alpha)}
$$

Where $\alpha$ is the smoothing coefficient (usually 1) and is used to avoid 
the zero-probability problem. It is often used to analyze categorical 
characteristic data, such as region, color, occupation type, etc.

#### Bernoulli Naive Bayes

Bernoulli naive Bayes is used when the features take only two values, such as 
0/1, True/False. It assumes that each feature follows a **Bernoulli 
Distribution** under the category $Y$:

$$
P(x_i|Y) = P_i^{x_i} (1-P_i)^{(1-x_i)},
$$

Where $P_i = P(x_i=1|Y)$. Often used in "word presence or absence" models for 
text classification (e.g., spam detection). Compared with the polynomial model, 
the Bernoulli model pays more attention to "whether a certain feature is 
included" rather than "the number of occurrences of features".

#### Multinomial Naive Bayes

Polynomial naive Bayes is often used for text analysis or document 
classification tasks. It assumes that features are represented as discrete 
frequencies or counts such as word frequency TF. Conditional probabilities are 
calculated as follows:

$$
P(X|Y) = \frac{(\sum_i x_i)! }{\prod_i x_i! } \prod_i P(x_i|Y)^{x_i},
$$

Where $x_i$ is the number of occurrences of the $i$word in the document. This 
model is particularly suitable for "Bag-of-Words" models, such as news 
classification, comment sentiment analysis, etc.

#### Complement Naive Bayes

Complement NB is an improvement of the polynomial model, It is mainly used for 
**class imbalanced** datasets [@zhang2004complement]. It estimates conditional 
probabilities by considering "samples outside of one category", reducing the 
bias of the model by mainstream categories:

$$
P(x_i|\bar{Y}) = \frac{\text{count}(x_i,\bar{Y}) + \alpha}
{\sum_{x_i'}(\text{count}(x_i',\bar{Y}) + \alpha)}
$$

This method works well when there are large differences in category 
proportions, For example, it is often used in tasks such as "accident severity 
classification" and "fraud detection".

#### Model comparison and application scenario summary

| Model type | Feature type | Common applications | Distributional assumptions | Remarks |
|-------------|---------------|----------------------|-----------------------------|----------|
| **Gaussian NB** | Continuous variables | medical diagnosis, price prediction | Gaussian distribution | more sensitive to outliers |
| **Categorical NB** | Categorical variables | Questionnaire, demographic analysis | Categorical distribution | fit discrete features |
| **Bernoulli NB** | Binary variables | Spam classification, whether the text <br> contains keywords | Bernoulli distribution | Focus on "presence or absence" |
| **Multinomial NB** | Count / frequency | text classification, sentiment analysis | Multinomial distribution | Fit for high dimensional sparse data |
| **Complement NB** | Imbalanced sample | fraud detection, traffic accident prediction | Polynomial complementary distribution | Improved imbalanced problem |


### Naive Bayesian application: NYC Crash Data

In this section, we will use the accident data set provided in the course to 
demonstrate how naive Bayesian models can be used to predict whether an 
accident is a "serious accident".

The goal here is to classify an original traffic accident record into:

-   severe (severe = 1) : Someone is injured or killed
-   No severe (severe = 0) : no injuries or deaths

This is a typical real world task: we want to quickly determine the severity of 
an accident as soon as it occurs or is recorded, so that we can direct 
resources, do risk monitoring, and observe patterns.

#### Modeling objectives

The task we are going to complete is a **binary classification problem** :

-   **Response variable (Y)** : 'severe_flag' Definition rule: If 
'persons_injured \> 0' or 'persons_killed \> 0', it is considered a serious 
accident and marked as 1, otherwise it is 0.

-   **Independent variable (X)** : We will use some features that can be 
extracted/discretized from accident records such as 
`borough`,`zip_code`,`crash_datetime`, 
`vehicle_type_code_1`,`contributing_factor_vehicle_1`... Most of these features 
are categorical factors, which are suitable for conditional probability based 
modeling in naive Bayes.

-   Load packages

```{r, message=FALSE, warning=FALSE}
library(arrow)
library(dplyr)
library(lubridate)
library(e1071) # naiveBayes
library(caret)
library(forcats)
```

-   Read our data

```{r}
# 1.
df <- arrow::read_feather("data/nyc_crashes_cleaned.feather")
```

-   Create the target variable severe_flag (severe accident =1, otherwise =0)

```{r}
df <- df %>%
  mutate(
    number_of_persons_injured = ifelse(is.na(number_of_persons_injured), 0, 
number_of_persons_injured),
    number_of_persons_killed = ifelse(is.na(number_of_persons_killed), 0, 
number_of_persons_killed),
    severe_flag = ifelse(number_of_persons_injured > 0 | 
number_of_persons_killed > 0, 1, 0)
  )
```

-   Extract hours from time and divide periods

```{r}
df <- df %>%
  mutate(
    crash_datetime = ymd_hms(crash_datetime, quiet = TRUE),
    hour = hour(crash_datetime),
    time_category = case_when(
      hour >= 0 & hour < 6 ~ "Late Night",
      hour >= 6 & hour < 12 ~ "Morning",
      hour >= 12 & hour < 18 ~ "Afternoon",
      hour >= 18 & hour < 24 ~ "Evening",
      TRUE ~ "Unknown"
    )
  )
```

-   Select features

```{r}
# borough: What borough is it in
# zip_code/zip_filled: Location code information
# time_category: Approximate time period when the accident occurred
# contributing_factor_vehicle_1: The primary contributing_factor for vehicle 1
# vehicle_type_code_1: Type of vehicle involved

# Note: We do not directly use the latitude and longitude, street names, these 
# are high base text/continuous values, not suitable for direct feeding naive 
# Bayes for classification.

model_df <- df %>%
  select(
    borough,
    zip_filled,
    time_category,
    contributing_factor_vehicle_1,
    vehicle_type_code_1,
    severe_flag
  ) %>%
  # Handling missing rows
  tidyr::drop_na() %>%
  mutate(
    borough = as.factor(borough),
    zip_filled = as.factor(zip_filled),
    time_category = as.factor(time_category),
    contributing_factor_vehicle_1 = as.factor(contributing_factor_vehicle_1),
    vehicle_type_code_1 = as.factor(vehicle_type_code_1),
    severe_flag = as.factor(severe_flag)
  )
```

-   Split data into training/testing sets and Train the Naive Bayes model.

```{r}
set.seed(42)
idx <- createDataPartition(model_df$severe_flag, p = 0.7, list = FALSE)
train_data <- model_df[idx, ]
test_data <- model_df[-idx, ]

nb_model <- naiveBayes(
  severe_flag ~ .,
  data = train_data,
  laplace = 1
)
```

-   Make predictions and evaluate the model.

```{r}
pred_class <- predict(nb_model, newdata = test_data)
pred_prob <- predict(nb_model, newdata = test_data, type = "raw")

confusionMatrix(
  data = pred_class,
  reference = test_data$severe_flag,
  positive = "1" # Treat "1" as a serious accident
)
```

##### Interpretation of our model results

As can be seen from the above table, the overall accuracy of the Naive Bayes 
model is about **60.6%**, which is higher than that of random guess, indicating 
that the model has a certain predictive ability in the overall classification.

However, the **Kappa value was 0.19**, indicating that the agreement between 
the model prediction results and the true label was weak.

In terms of classification ability, the model identified "non-serious 
accidents" better (specificity 0.77). However, the ability to identify "serious 
accidents" was relatively weak (sensitivity 0.43), that is, there was a certain 
proportion of missed detection.

The main reasons for this may include:

-   **Imbalanced Data** : The samples of non-serious accidents are far more 
than those of serious accidents, resulting in the model being biased towards 
the majority class.

-   **Naive Assumption** : Naive Bayes assumes that each feature is independent 
of each other, but this assumption is often not true in real traffic accident 
data.

-   **Limited feature information** : The variables used (such as time period, 
vehicle type, causative factors, etc.) are relatively macroscopic and lack 
continuity or quantitative characteristics, which limits the accuracy of the 
model.

Nevertheless, the naive Bayesian model has the following advantages:

-   fast training and prediction.
-   strong interpretation of parameters
-   Suitable as a baseline model (baseline) for more complex models (e.g. 
Random Forest, XGBoost).

Therefore, the results of this model can be used as a reference basis for 
subsequent improvements, and further improvements in model performance are 
expected after the introduction of more features or balanced samples.